---
title: "Large-scale clustering in Bioconductor"
author: "Davide Risso and Stephanie Hicks"
date: "Last modified: November 18, 2020; Compiled: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmarkdown::html_document:
   highlight: pygments
   toc: true
   toc_depth: 3
   fig_width: 5
vignette: >
  %\VignetteIndexEntry{CZI Seed Network Meeting 2020 Demo}
  %\VignetteEncoding[ut8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this demo, we will showcase an end-to-end clustering pipeline, starting from the count data matrix stored in HDF5 (similar to what one would download from the HCA data portal) all the way to visualization and interpretation of the clustering results.

While we use a small-ish dataset for this demo for convenience, the code is computationally efficient even for (very) large datasets.

```{r packages}
library(SingleCellExperiment)
library(TENxPBMCData)
library(scater)
library(scran)
library(scry)
library(NewWave)
library(mbkmeans)
```

## Getting the data

```{r data}
sce <- TENxPBMCData("pbmc4k")
sce
counts(sce)
seed(counts(sce))
```

## Filtering and normalization

### Removing low-quality cells

First, we use the `scater` package to compute a set of 
QC measures and filter out the low-quality samples.

Here, we exclude those cells that have a too high percentage of mitochondrial genes or for which we detect too few genes.

```{r filter}
sce <- addPerCellQC(sce, 
            subsets = list(Mito = grep("^MT-", rowData(sce)$Symbol_TENx)))
high_mito <- isOutlier(sce$subsets_Mito_percent, 
                       nmads = 3, type="higher")
low_detection <- (sce$detected < 1000)
high_counts <- sce$sum > 45000
sce <- sce[,!high_mito & !low_detection & !high_counts]
sce
```

### Removing lowly expressed genes

Next, we remove the lowly expressed genes. Here, 
we keep only those genes that have at least 1 UMI 
in at least 5% of the data. These threshold are
dataset-specific and may need to be taylored to 
specific applications.

```{r qc-gene-filter}
num_reads <- 1
num_cells <- 0.01*ncol(sce)
keep <- which(DelayedArray::rowSums(counts(sce) >= num_reads ) >= num_cells)
sce <- sce[keep,]
sce
```

These leaves us with `length(keep)` genes.

### Normalization

Here, we apply `mbkmeans` (`k=10` and batch size of 500) as a preliminary step to `scran` normalization.

```{r mbkmeans_full}
set.seed(18)
mbk <- mbkmeans(sce, whichAssay = "counts", reduceMethod = NA,
                  clusters=10, batch_size = 500)
sce$mbk10 <- paste0("mbk", mbk$Clusters)
table(mbk$Clusters)
```

We then compute the normalization factors and normalize the data.

```{r scran}
sce <- computeSumFactors(sce, cluster=mbk$Clusters, min.mean=0.1)
sce <- logNormCounts(sce)
sce
```

## Dimensionality reduction

### PCA on normalized values

Here, we compute the first 50 principal components using the top variable genes.

```{r pca}
sce <- scater::runPCA(sce, ncomponents = 50,
                      ntop = 1000,
                      scale = TRUE,
                      BSPARAM = BiocSingular::RandomParam())
plotPCA(sce, colour_by = "mbk10")
```

### GLM-PCA

```{r glmpca}
sce <- nullResiduals(sce, assay="counts", type="deviance")
sce <- scater::runPCA(sce, ncomponents = 50,
                      ntop = 1000,
                      exprs_values = "binomial_deviance_residuals",
                      scale = TRUE, name = "GLM-PCA",
                      BSPARAM = BiocSingular::RandomParam())
plotReducedDim(sce, dimred = "GLM-PCA", colour_by = "mbk10")
```

### ZINB-WaVE

```{r newwave, eval=FALSE}
dec <- modelGeneVar(sce)
hvg <- getTopHVGs(dec, n=1000)
sce_red <- sce[hvg,]

sce_red <- newWave(sce_red, K=2,
               n_gene_disp = 100,
               n_gene_par = 100, 
               n_cell_par = 100)
plotReducedDim(sce_red, dimred = "NewWave", colour_by = "mbk10")
```

## Clustering

Here, we use the GLM-PCA results to obtain the final cluster labels. We use two alternative approaches: Louvain and mini-batch k-means.

```{r louvain}
g <- buildSNNGraph(sce, k=10, use.dimred = "GLM-PCA")
lou <- igraph::cluster_louvain(g)
sce$louvain <- paste0("Louvain", lou$membership)
table(sce$louvain)
```

```{r select_k}
k_list <- seq(5, 20)
km_res <- lapply(k_list, function(k) {
    mbkmeans(sce, clusters = k, 
             batch_size = 500,
             reduceMethod = "GLM-PCA",
             calc_wcss = TRUE)
})
wcss <- sapply(km_res, function(x) sum(x$WCSS_per_cluster))
plot(k_list, wcss, type = "b")
```

```{r minibatch}
sce$kmeans <- paste0("mbk", km_res[[which(k_list==13)]]$Clusters)
table(sce$kmeans)
table(sce$kmeans, sce$louvain)

plotReducedDim(sce, dimred = "GLM-PCA", colour_by = "louvain")
plotReducedDim(sce, dimred = "GLM-PCA", colour_by = "kmeans")
```

## Cluster interpretation


